Hybrid MILP to efficiently and accuratly \\ solve hard DNN verification instances

### 2024-09-20 ICLR25/experiments.tex
table
formula

\begin{table}[t!]	
	\centering
	\begin{tabular}{|c||c|c|c|c|}
		\hline
		\text{Number $K$}  &  \text{Without ratio} &\text{With ratio} &  \text{Incoming Weights} \cite{ DivideAndSlide} & \text{Random} \\ \hline
		\hline
		0  &   1.761  & 1.761 & 1.761 & 1.761  \\ \hline
		5  & 1.532  & 1.533 & &   \\ \hline
		10  & 1.371 & 1.373 & 1.651 & 1.696  \\ \hline
		15  & 1.247 & 1.244 &   &   \\ \hline
		20  & 1.145 & 1.136 & 1.557 & 1.619  \\ \hline
		25  & 1.061 & 1.057 &  &   \\ \hline
		30  & 0.989 & 1.067 & 1.489 & 1.546  \\ \hline
		35  & 0.934 & 1.091 &  &   \\ \hline
		40  &  & & 1.447 & 1.469  \\ \hline
		50  &  & & 1.426 & 1.388  \\ \hline
		116 (max) &  0.895 & 0.895 & 1.424 (max=$61$) & 0.895  \\ \hline
	\end{tabular}
	\caption{Average uncertainty of $\MILP_Z$ for nodes of the third layer, for $Z$ with $K$ ReLU nodes of the 1st and 2nd layer (compensating strength vs \cite{ DivideAndSlide} vs random choice).}
	\label{tab:example1}
	\vspace{-0.6cm}
\end{table}

>>>>

\begin{table}[t!]	
	\centering
	\begin{tabular}{|c||c|c|c|}
		\hline
		\text{Number $K$}  &  \bf Utility&  \text{Incoming Weights} \cite{ DivideAndSlide} & \text{Random} \\ \hline
		\hline
		0  &   1.761  & 1.761 & 1.761  \\ \hline
		5  & \bf 1.532  & &   \\ \hline
		10  & \bf 1.371  & 1.651 & 1.696  \\ \hline
		15  & \bf 1.247  &   &   \\ \hline
		20  & \bf 1.145  & 1.557 & 1.619  \\ \hline
		25  & \bf 1.061  &  &   \\ \hline
		30  & \bf 0.989  & 1.489 & 1.546  \\ \hline
		35  & \bf 0.934  &  &   \\ \hline
		40  & \bf 0.914 & 1.447 & 1.469  \\ \hline
		116 (max) &  0.895 & 1.424 (max=$61$) & 0.895  \\ \hline
	\end{tabular}
	\caption{Average uncertainty of $\MILP_Z$ for nodes of the third layer, for $Z$ with $K$ ReLU nodes of the 1st and 2nd layer (utility function vs \cite{ DivideAndSlide} vs random choice).}
	\label{tab:example1}
	\vspace{-0.6cm}
\end{table}

### 2024-09-20 ICLR25/formula.tex
formula

Let $z$ be a neuron of the DNN we want to evaluate, for instance computing an upper bound on the value it can take. We denote by $n< z$ the fact that a neuron appears on a layer before the layer of $z$, and $n \leq z$ if it can also be $z$.
Let us denote (Sol\_max$_X^z(n))_{n \leq z})$ is a solution of $\mathcal{M}_X$ 
maximizing $z$. In particular, Sol\_max$_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider (Sol\_max$_\emptyset^z(n))_{n \leq z})$ (this is obtained by {\em one} call to an LP solver, so can be obtained very efficiently).
An accurate utility function is to evaluate 
Improve\_max$^z(a)=$ Sol\_max$_\emptyset^z(z) -$ Sol\_max$_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
Similarly, we define (Sol\_min$_\emptyset^z(n))_{n \leq z})$ and 
Improve\_min$_a^z$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.

Instead, we focus on the following upper bounding of Improve\_max$_a^z$, thanks to Proposition \ref{LP}.

@@

\subsubsection*{Compute the improvement of a single neuron}
\fi

@@

%For one layer before the target layer, the formula is simple and most accurate. 
Suppose $a$ is a neuron in one layer before $z$. To estimate Improve\_max$_a^z$:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

\begin{definition}
	For all neurons $n$, let $sol(n)=$Sol\_max$_\emptyset^z(n)$ be the value of neuron $a$
	in the solution of the LP instance Sol\_max$_\emptyset^z$ to maximize $z$.

	We define : 
	\begin{align}
		Utility\_max^z(a) = W_{az} \times (sol(\hat{a})- \ReLU(sol(a))).
	\end{align}

@@

\begin{proof}
Indeed, consider $(sol'(n))_{n \leq z})_{n \leq z})$ with
$sol'(n)=sol(n)$ for all $n \notin \{z,\hat{a}\}$.
In particular,  $sol'(a) = sol(a)$.

@@

@@

\end{definition}

@@

\subsubsection*{Two Layer before $z$}

Suppose $a$ is a neuron in two layers before $z$, we use $b$ to denote neurons in the layer between $a$ and $z$.

In this case, the formula will be more complex and less accurate. We will still focus on the source $sol(\hat{a})-\ReLU(sol(a))$. Basically, we hope to use $\sum_b W_{bz}W_{ab}$ to replace the coefficient $W_{az}$ in last subsection. However, directly using $\sum_b W_{bz}W_{ab}$ will lose too much accuracy. So we try to improve the accuracy by considering more details from the LP solutions.

To estimate Improve\_max$_a^z$ , we consider the following help function:

\begin{definition} \label{2layer}
	Let $\UB,\LB$ be the function of precomputed upper and lower bound before the layer of $z$.
	We define the following function $k$ for neurons $b$ in one layer before $z$:
	\begin{align}
		k(b) =
		\begin{cases}
			1, & \text{if } sol(b) > 0 \text{ and } W_{bz} < 0\\
			0, & \text{if } sol(b) \leq 0 \text{ and } W_{bz} < 0\\
			\frac{\UB(b)}{\UB(b)-\LB(b)}, & \text{if }W_{bz} > 0
		\end{cases}
	\end{align} 
\end{definition}

We may use the following formula to estimate the improvement:  \begin{align*}
	Improve\_max^z(a) \approx (sol(\hat{a})-\ReLU(sol(a)))\sum_b W_{bz}W_{ab}k(b).
\end{align*}However, the error is still too large. This is because the improved value of a neuron $b$ can not exceed its precomputed upper and lower bound. Considering this, we can update our formula.

\begin{definition}
	Continue the assumption in Definition \ref{2layer}. To compute the improvement of of upper bound of $z$ by $a$, we define a distance function $D$:
	\begin{align}
		&D(a) = \ReLU(sol(a))-sol(\hat{a})\\
			&D(b) =
		\begin{cases}
			\max(\LB(b), 0, sol(b)+D(a)k(b)W_{ab})-sol(\hat{b}), & \text{if }W_{bz} < 0\\
			\min(\UB(b), sol(\hat{b})+D(a)k(b)W_{ab})-sol(\hat{b}), & \text{if }  W_{bz} > 0
		\end{cases}\\
		&D(z) = \sum_b D(b)W_{bz}\\
		&Utility\_max^z(a) = -D(z)
	\end{align}
\end{definition}
 This is the formula to compute improvement of neurons two layers before.

\subsubsection*{Three Layer before $z$}

>>>>

Let $z$ be a neuron of the DNN we want to evaluate, for instance computing an upper bound on the value it can take. We denote by $n< z$ the fact that a neuron appears on a layer before the layer of $z$, and $n \leq z$ if it can also be $z$.
Let us denote (Sol\_max$_X^z(n))_{n \leq z}$ is a solution of $\mathcal{M}_X$ 
maximizing $z$. In particular, Sol\_max$_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider (Sol\_max$_\emptyset^z(n))_{n \leq z}$ (this is obtained by {\em one} call to an LP solver, so can be obtained very efficiently).
An accurate utility function is to evaluate 
Improve\_max$^z(a)=$ Sol\_max$_\emptyset^z(z) -$ Sol\_max$_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have Improve\_max$^z(a)\geq 0$ as Sol\_max$_{\{a\}}^z$ fullfils all the constraints of 
$\mathcal{M}_\emptyset$, so Sol\_max$_{\{a\}}^z(z) \leq$ Sol\_max$_\emptyset^z(z)$.

Similarly, we define (Sol\_min$_\emptyset^z(n))_{n \leq z}$ and 
Improve\_min$_a^z$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
Instead, we focus on the following upper bounding of Improve\_max$_a^z$.

@@

\subsubsection*{Compute the improvement of a single neuron}

@@

\fi

For all neurons $n$, let $sol(n)=$Sol\_max$_\emptyset^z(n)$ be the value of neuron $a$
in the solution of the LP instance Sol\_max$_\emptyset^z$ to maximize $z$.

%For one layer before the target layer, the formula is simple and most accurate. 
To estimate Improve\_max$_a^z$, we define Utility\_max$^z(a)$, first for neurons $a$ one layer before $z$:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

	\begin{align}
		Utility\_max^z(a) = |W_{az} \times (sol(\hat{a})- \ReLU(sol(a)))|.
	\end{align}

@@

\begin{proof}
Consider $(sol'(n))_{n \leq z})_{n \leq z})$ with
$sol'(n)=sol(n)$ for all $n \notin \{z,\hat{a}\}$.
In particular,  $sol'(a) = sol(a)$.

@@

Thus, Utility\_max$^z(a)$ can be used to approximate 
Improve\_max$^z(a)$. In particular, consider a very frequent case where $W_{az}<0$.
In this case, to maximize $z$, the LP engine sets $sol(\hat{a})$ to its minimal value\dots which happens to be $sol(\hat{a})=ReLU(a)$, thanks to Proposition \ref{LP}. 
In this frequent case, we have Utility\_max$^z(a)=0$, and thus Improve\_max$^z(a)=0$. 
In particular, this node will have the smallest utility (thus will not get picked in the open nodes $X$), and indeed it is not having any impact on  Sol\_max$_{\{a\}}^z(z)$. 
This is one stricking difference (but not the only one) with choosing utility based on 
$|W_{az}|$ \cite{DivideAndSlide}.

@@

@@

\medskip

Now, consider neurons $a$ two layers before $z$. We use $b$ to denote neurons in the layer between $a$ and $z$. If we open $a$, then the change in the weight of $a$ will still be $sol(\hat{a})-\ReLU(sol(a))$ as above. Its impact on $z$ is no more direct with $W_{az}$, but it is through all the $b$ in the intermediate layer. 
Based on Proposition \ref{LP}, we can evaluate how opening $a$ impacts each value of 
$\hat{b}$, through coefficient $k(\hat{b})$, by using the upper and lower bound
$\UB(b),\LB(b)$:
\begin{itemize}
 \item If $W_{bz}\leq0$, then $k(\hat{b})=W_{ab}$ if $sol(b) > 0$, and $k(\hat{b})=0$ if $sol(b) < 0$
 \item If $W_{bz}>0$, then $k(\hat{b})=W_{ab} \frac{\UB(b)}{\UB(b)-\LB(b)}$
\end{itemize}
%$\sum_b W_{bz}W_{ab}$ to replace the coefficient $W_{az}$ in last subsection.

%To estimate Improve\_max$_a^z$ , 
Indeed, if $W_{bz}\leq0$, then the LP solver sets $sol(\hat{b})$ to the lowest possible value to maximize $z$, which happens to be $ReLU(b)$ according to Proposition \ref{LP}.
If $sol(b) < 0$, then we have a clip at $sol(\hat{b})=ReLU(b)=0$ and opening $a$ does not help. If $sol(b) > 0$, then we have $sol(\hat{b})=ReLU(b)=b$ and every change from 
$a$ is reflected with a coefficient $W_{ab}$ on $b$. 
Last, if $W_{bz}>0$, then according to Proposition \ref{LP}, the LP solver
sets $sol(\hat{b}) = sol(b) \frac{\UB(b)}{\UB(b)-\LB(b)}$ to maximize $z$.

%We then define 
%\begin{align*}
%	Utility\_max^z(a) = (sol(\hat{a})-\ReLU(sol(a)))\sum_b k(b).
%\end{align*}

We define a distance $D$ to account for the fact that a neuron $b$ 
can not exceed $\LB(b),\UB(b)$:
	\begin{align*}
		&D(\hat{a}) = sol(\hat{a}) - \ReLU(sol(a)) \geq 0\\
			&D(\hat{b}) =
		\begin{cases}
			sol(\hat{b}) - \min(\UB(b), sol(\hat{b}) - D(\hat{a}) k(\hat{b})), & \text{if }  W_{bz} > 0 \\
			sol(\hat{b}) - \max(\LB(b), 0, sol(\hat{b}) -D(\hat{a}) k(\hat{b})), & \text{if }W_{bz} < 0\\
		\end{cases}
		\end{align*}

We can now set

$$ Utility\_max^z(a) = \sum_b W_{bz} D(\hat{b})$$

We proceed inductively in the same way to define Utility$\_max^z(a)$ for deeper neuron $a$.

\iffalse
\subsubsection*{Three Layer before  $z$}
